Temporal trends (NAMCS)

### 2017-07-24 ms1/ms1_display.Rmd
streamlining

pctize = function(df, denom_name) {
  for (n_name in Filter(function(x) str_detect(x, '^n_'), names(df))) {
    pct_name = str_replace(n_name, '^n_', 'pct_')
    df[[pct_name]] = df[[n_name]] / df[[denom_name]] * 100
  }
  df
}

trend_p = function(x) t.test(diff(x))$p.value
```

@@

```{r}
model_ci = function(df) {
  mutate(df, ci_range=1.96*std.error,
         ci_low=estimate-ci_range,
         ci_high=estimate+ci_range)
}

@@

```{r}
totals_dat = read_tsv('tables/tbl_cohort_totals.tsv') %>%
  pctize('n_bene') %>%
  select(year, n_bene, mean_age, sd_age, mean_n_cc, sd_n_cc,
         pct_female, pct_white, pct_dual,
         pct_region_south, pct_region_midwest, pct_region_northeast, pct_region_west)

@@

```{r}
linear_trend_p = function(x) tidy(t.test(diff(x)))$p.value

totals_dat %>%
  gather('var', 'val', -year) %>%
  arrange(var, year, val) %>%
  group_by(var) %>%
  summarize(p=linear_trend_p(val)) %>%
  mutate(adj.p=p.adjust(p, method='BH'),
         sig=if_else(adj.p<0.05, '*', '')) %>%
  kable
```

@@

```{r}
cohort_dat = read_tsv('tables/tbl_cohort_claims_by_abx.tsv')

top_abx = cohort_dat %>%
  group_by(antibiotic) %>%
  summarize(cpkp=mean(cpkp)) %>%
  arrange(desc(cpkp)) %$%
  head(antibiotic, 10)

@@

```{r}
cohort_dat %>%
  top_abx_f() %>%
  ggplot(aes(x=year, y=cpkp)) +
  facet_wrap(~ antibiotic, ncol=3, scales='free_y') +
  geom_point() + geom_line() +
  theme_minimal() +
  ggtitle('Unadjusted consumption, in cohort')

cohort_dat %>%
  top_abx_f() %>%
  filter(year %in% c(2011, 2014)) %>%
  spread(year, cpkp) %>%
  mutate(pct_change=(`2014`-`2011`)/`2011`*100) %>%
  kable
```

P-values for trends in the data. (These are $t$-tests on the differences between subsequent years, comparing their mean to zero.)

```{r}
cohort_dat %>%
  top_abx_f() %>%
  arrange(antibiotic, year) %>%
  group_by(antibiotic) %>%
  summarize(p.value=trend_p(cpkp)) %>%
  ungroup() %>%
  mutate(adj.p.value=p.adjust(p.value, method='BH')) %>%
  select(antibiotic, adj.p.value) %>%
  mutate(sig=if_else(adj.p.value < 0.05, '*', '')) %>%
  kable
```

@@

```{r claims_by_age}
read_tsv('tables/tbl_cohort_claims_by_age.tsv') %>%
  mutate(cpkp=round(cpkp)) %>%
  select(year, age_group, cpkp) %>%
  spread(year, cpkp) %>%
  kable
```

@@

```{r}
read_tsv('tables/tbl_cohort_claims_by_sex.tsv') %>%
  select(sex, year, cpkp) %>%
  spread(year, cpkp) %>%
  arrange(desc(`2011`)) %>% 
  mutate_at(vars(-sex), round) %>%
  kable
```

@@

```{r}
read_tsv('tables/tbl_cohort_claims_by_region.tsv') %>%
  select(region, year, cpkp) %>%
  spread(year, cpkp) %>%
  arrange(desc(`2011`)) %>% 
  mutate_at(vars(-region), round) %>%
  kable
```

@@

```{r}
read_tsv('tables/tbl_cohort_claims_by_race.tsv') %>%
  select(race, year, cpkp) %>%
  spread(year, cpkp) %>%
  arrange(desc(`2011`)) %>% 
  mutate_at(vars(-race), round) %>%
  kable
```

@@

"Adjusted" means the `year` coefficient in `y ~ year + age + n_cc + is_female + is_dual + is_white + region`.

```{r adjusted_changes, cache.extra=md5sum('tables/tbl_cohort_model_abx.tsv')}
x = read_tsv('tables/tbl_cohort_model_abx.tsv') %>%
  filter(term=='year', model %in% top_abx) %>%
  arrange(estimate) %>%
  mutate(cpkp_per_year=estimate*1000,
         antibiotic=as_factor(model)) %>%
  select(antibiotic, cpkp_per_year)

p = x %>%
  ggplot(aes(x=antibiotic, y=cpkp_per_year)) +
  geom_point() +
  geom_linerange(aes(x=antibiotic, ymin=0, ymax=cpkp_per_year)) +
  coord_flip() +
  ggtitle('Adjusted changes in consumption') +
  xlab('') +
  ylab('adjusted yearly change in claims per 1,000 beneficiaries per year') +
  scale_y_continuous(limits=c(-16, 17), breaks=c(-15, -10, -5, 0, 5, 10, 15)) +
  theme_minimal()

show(p)
ggsave('fig/fig_adjusted_changes.pdf', plot=p)

>>>>

pctize = function(df, denom_name) {
  for (n_name in Filter(function(x) str_detect(x, '^n_'), names(df))) {
    pct_name = str_replace(n_name, '^n_', 'pct_')
    df[[pct_name]] = df[[n_name]] / df[[denom_name]] * 100
  }
  df
}
```

@@

```{r, eval=FALSE}
model_ci = function(df) {
  mutate(df, ci_range=1.96*std.error,
         ci_low=estimate-ci_range,
         ci_high=estimate+ci_range)
}

@@

```{r}
totals_dat = read_tsv('tables/totals.tsv') %>%
  pctize('n_bene') %>%
  select(year, n_bene, mean_age, sd_age, mean_n_cc, sd_n_cc,
         pct_female, pct_white, pct_dual,
         pct_region_south, pct_region_midwest, pct_region_northeast, pct_region_west)

@@

```{r}
trend_p = function(x) tidy(t.test(diff(x)))$p.value

totals_dat %>%
  gather('var', 'val', -year) %>%
  arrange(var, year, val) %>%
  group_by(var) %>%
  summarize(p=trend_p(val)) %>%
  mutate(adj.p=p.adjust(p, method='BH'),
         sig=if_else(adj.p<0.05, '*', '')) %>%
  kable
```

@@

```{r}
claims_by_abx = read_tsv('tables/claims_by_abx.tsv')

top_abx = claims_by_abx %>%
  group_by(antibiotic) %>%
  summarize(cpkp=mean(cpkp)) %>%
  arrange(desc(cpkp)) %$%
  head(antibiotic, 10)

@@

```{r}
claims_by_abx %>%
  top_abx_f() %>%
  ggplot(aes(x=year, y=cpkp)) +
  facet_wrap(~ antibiotic, ncol=3, scales='free_y') +
  geom_point() + geom_line() +
  theme_minimal() +
  ggtitle('Unadjusted consumption')

claims_by_abx %>%
  filter(antibiotic %in% top_abx) %>%
  mutate(antibiotic=factor(antibiotic, levels=top_abx)) %>%
  ggplot(aes(x=year, y=cpkp)) +
  facet_wrap(~ antibiotic, ncol=3) +
  geom_point() + geom_line() +
  theme_minimal() +
  ggtitle('Unadjusted consumption')

claims_by_abx %>%
  top_abx_f() %>%
  filter(year %in% c(2011, 2014)) %>%
  spread(year, cpkp) %>%
  mutate(pct_change=(`2014`-`2011`)/`2011`*100) %>%
  kable
```

@@

```{r claims_by_age}
read_tsv('tables/claims_by_age_group.tsv') %>%
  mutate(cpkp=round(cpkp)) %>%
  select(year, age_group, cpkp) %>%
  spread(year, cpkp) %>%
  kable
```

@@

```{r}
read_tsv('tables/claims_by_sex.tsv') %>%
  select(sex, year, cpkp) %>%
  spread(year, cpkp) %>%
  arrange(desc(`2011`)) %>% 
  mutate_at(vars(-sex), round) %>%
  kable
```

@@

```{r}
read_tsv('tables/claims_by_region.tsv') %>%
  select(region, year, cpkp) %>%
  spread(year, cpkp) %>%
  arrange(desc(`2011`)) %>% 
  mutate_at(vars(-region), round) %>%
  kable
```

@@

```{r}
read_tsv('tables/claims_by_race.tsv') %>%
  select(race, year, cpkp) %>%
  spread(year, cpkp) %>%
  arrange(desc(`2011`)) %>% 
  mutate_at(vars(-race), round) %>%
  kable
```

@@

"Adjusted" means the `year` coefficient in `y ~ year + age + n_cc + is_dual + region`. Gender and race, though important covariates, are not confounded with time.

```{r adjusted_changes, cache.extra=md5sum('tables/model_abx.tsv')}
x = read_tsv('tables/model_abx.tsv') %>%
  filter(model=='poisson', term=='year', antibiotic %in% top_abx) %>%
  arrange(estimate) %>%
  mutate(ci=1.96*std.error,
         pct_change=exp(estimate)-1,
         ci_low=exp(estimate-ci)-1,
         ci_high=exp(estimate+ci)-1) %>%
  select(antibiotic, pct_change, ci_low, ci_high) %>%
  arrange(pct_change) %>% mutate(antibiotic=forcats::fct_inorder(antibiotic)) %>%
  mutate_if(is.numeric, function(x) x*100)

p = x %>%
  ggplot(aes(x=antibiotic, y=pct_change, ymin=ci_low, ymax=ci_high)) +
  geom_point() +
  #geom_linerange(aes(x=antibiotic, ymin=0, ymax=cpkp_per_year)) +
  geom_linerange() +
  coord_flip() +
  ggtitle('Adjusted changes in consumption') +
  xlab('') +
  ylab('adjusted yearly change in claims per beneficiary per year') +
  #scale_y_continuous(limits=c(-16, 17), breaks=c(-15, -10, -5, 0, 5, 10, 15)) +
  theme_minimal()

show(p)
#ggsave('fig/fig_adjusted_changes.pdf', plot=p)

### 2017-07-24 ms1/namcs_trends.Rmd
streamlining

top_abx = read_tsv('../../db/namcs/analysis/top_abx.tsv')
abx_cat = read_tsv('../../db/namcs/data/parse/drug_codes/abx_codes.tsv')

@@

abx_visit = left_join(top_abx, drugs, by='desc') %>%
  count(visit_id, abx) %>% ungroup() %>%
  spread(abx, n)

@@

I consider the same drugs as last time:

```{r}
kable(top_abx)
```

@@

But the number of Americans went up (but the number of children went down):

@@

```{r overall_model1}
svyglm(n_abx ~ year, design=design, family=quasipoisson(link='identity')) %>%
  tidy %>% kable
```

@@

```{r overall_model2}
overall_trend %>%
  mutate(year=year-min(year)) %>%
  lm(abx_per_capita ~ year, weights=1.0/se_per_capita^2, data=.) %>%
  tidy %>% kable
```

@@

```{r drugs_time}
# apc = antibiotic per capita
f = function(design, d) {
  svyby(as.formula(paste0('~', d)), ~year, design=design, svytotal) %>%
    as_data_frame() %>%
    rename_(.dots=setNames(d, 'n_abx')) %>%
    mutate(abx=d)
}

g = function(design) {
  lapply(top_abx$abx, function(x) f(design, x)) %>% bind_rows() %>%
    left_join(census, by='year') %>%
    mutate(apc=n_abx/n_total, se=se/n_total)
}

abx_time = g(design)
abx_time_adult = g(adult_design)
```

```{r drugs_time_plot}
abx_time %>%
  mutate(ci=1.96*se, ymin=apc-ci, ymax=apc+ci) %>%
  ggplot(aes(x=year, y=apc, ymin=ymin, ymax=ymax)) +
  facet_wrap(~abx, ncol=3, scales='free_y') +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()
```

@@

It doesn't look by eye that any trends are particularly strong.

I used weighted linear regression to look for trends in totals:

```{r drugs_time_totals_model}
abx_total_models = abx_time %>%
  group_by(abx) %>%
  do(tidy(lm(apc ~ year, weight=1.0/se^2, data=.))) %>%
  ungroup()

abx_total_models %>%
  filter(term=='year') %>% select(-term) %>%
  arrange(estimate) %>%
  kable

abx_total_models %>%
  filter(term=='year') %>% select(-term) %>%
  mutate(ci=1.96*std.error, ymin=estimate-ci, ymax=estimate+ci) %>%
  mutate(abx=forcats::fct_reorder(abx, estimate)) %>%
  ggplot(aes(x=abx, y=estimate, ymin=ymin, ymax=ymax)) +
  geom_point() + geom_linerange() +
  coord_flip() +
  theme_minimal()
```

I used Poisson regressions to see if anything was changing *per visit* with time. Only levo:

```{r drugs_time_model}
qpglm = function(design, frmla) {
  svyglm(frmla, design=design, family=quasipoisson(link='identity'))
}

f = function(design, d) {
  qpglm(design, as.formula(paste0(d, '~ year'))) %>%
    tidy %>%
    mutate(abx=d) %>%
    filter(term != '(Intercept)')
}

g = function(design) lapply(top_abx$abx, function(x) f(design, x)) %>% bind_rows()

abx_time_model = g(design)
abx_time_model_adult = g(adult_design)
```

```{r drugs_time_model_table}
abx_time_model %>%
  select(abx, estimate, p.value) %>%
  mutate(adj.p.value=p.adjust(p.value, method='BH')) %>%
  arrange(p.value) %>%
  kable
```

### Levo trend

```{r levo_ratio}
levo_time = abx_time %>%
  filter(abx=='levo') %>%
  arrange(year)

t1 = levo_time %>% filter(year==2011) %$% apc
v1 = levo_time %>% filter(year==2011) %$% se
t2 = levo_time %>% filter(year==2014) %$% apc
v2 = levo_time %>% filter(year==2014) %$% se

tr = t2 / t1
vr = v1 / t2^2 + t1^1 / t2^4 * v2
sdr = sqrt(vr)

show_tot = function(tot, v) sprintf('%g (±%g)', round(tot, 4), round(v, 4))
show_pct = function(tot, v) sprintf('%g%% (±%g%%)', round((tot - 1)*100, digits=1), round(sqrt(v)*100, digits=1))
```

Levo increased from `r show_tot(t1, v1)` to `r show_tot(t2, v2)` per capita, an increase of `r show_pct(tr, vr)`.

# Conclusions

Levo is definitely going up.

>>>>

top_abx = read_tsv('namcs_top_abx.tsv')
abx_cat = read_tsv('../../db/namcs/data/parse/drug_codes/abx_codes.tsv')

@@

abx_visit = top_abx %>%
  select(desc, abx) %>%
  left_join(drugs, by='desc') %>%
  count(visit_id, abx) %>% ungroup() %>%
  spread(abx, n)

@@

I consider the same drugs as in Medicare, except for clindamycin, which I drop,
because NAMCS doesn't have enough information to distinguish between oral and
topical formulations.

```{r top_abx_table}
top_abx %>%
  select(medicare_abx, abx) %>%
  kable
```

@@

However, the number of Americans went up (but the number of children went down):

@@

```{r lm_functions}
# survey poisson model
svy_pm = function(design, frmla) svyglm(frmla, design=design, family='poisson') %>% tidy

# log glm
log_glm = function(df, frmla, weights=NULL) glm(formula=as.formula(frmla), data=df, family=gaussian(link='log'), weights=weights) %>% tidy
```

```{r overall_per_visit}
svy_pm(design, n_abx ~ year) %>% kable
```

@@

```{r overall_per_capita}
overall_trend %>%
  mutate(year=year-min(year)) %>%
  log_glm('abx_per_capita ~ year', weights=1.0/.$se_per_capita^2) %>%
  kable
```

@@

#### Per capita

```{r drugs_per_capita_model}
n_abx_f = function(design, a) {
  svyby(as.formula(paste0('~', a)), ~year, design=design, svytotal) %>%
    as_data_frame() %>%
    rename_(.dots=setNames(a, 'n_abx')) %>%
    mutate(abx=a)
}

# apc = antibiotic per capita
apc_f = function(design, denom) {
  lapply(top_abx$abx, function(x) n_abx_f(design, x)) %>% bind_rows() %>%
    left_join(census, by='year') %>%
    rename_(.dots=setNames(denom, 'denom')) %>%
    mutate(apc=n_abx/denom, se=se/denom)
}

abx_time = apc_f(design, 'n_total')
abx_time_adult = apc_f(adult_design, 'n_adult')
```

```{r drugs_per_capita_time}
abx_time %>%
  mutate(ci=1.96*se, ymin=apc-ci, ymax=apc+ci) %>%
  ggplot(aes(x=year, y=apc, ymin=ymin, ymax=ymax)) +
  facet_wrap(~abx, ncol=3, scales='free_y') +
  geom_point() + geom_line() + geom_linerange() +
  theme_minimal()
```

@@

It doesn't look by eye that any trends are particularly strong. I used weighted
log regression to look for trends in totals:

```{r drugs_per_capita_time_trends}
abx_total_models = abx_time %>%
  group_by(abx) %>%
  do(log_glm(., 'apc ~ year', weights=1.0/.$se^2)) %>%
  ungroup()

abx_total_models %>%
  filter(term=='year') %>% select(-term) %>%
  arrange(estimate) %>%
  kable
```

```{r drugs_per_capita_figure}
bind_rows(
  read_tsv('tables/model_abx.tsv') %>%
    filter(term=='year', model=='poisson') %>%
    left_join(top_abx, by=c('antibiotic'='medicare_abx')) %>%
    mutate(abx=if_else(is.na(abx), antibiotic, abx)) %>%
    select(abx, estimate, std.error) %>%
    mutate(data_source='medicare'),
  abx_total_models %>%
    filter(term=='year') %>%
    select(abx, estimate, std.error) %>%
    mutate(data_source='namcs')
) %>%
  #mutate(abx=factor(abx, levels=top_abx$abx)) %>%
#abx_total_models %>%
#  filter(term=='year') %>%
  mutate(ci=1.96*std.error, y=exp(estimate)-1, ymin=exp(estimate-ci)-1, ymax=exp(estimate+ci)-1) %>%
  #select(abx, y, ymin, ymax) %>%
  mutate_if(is.numeric, function(x) x*100) %>%
  #arrange(y) %>% mutate(abx=forcats::fct_inorder(abx)) %>%
  #ggplot(aes(x=abx, y=y, ymin=ymin, ymax=ymax)) +
  ggplot(aes(x=abx, y=y, ymin=ymin, ymax=ymax, color=data_source)) +
  geom_linerange() +
  geom_point() +
  coord_flip() +
  ylab('pct change in consumption per capita') +
  theme_minimal()
```

#### Per visit

I used Poisson regressions to see if anything was changing *per visit* with
time: levo and amox/clav. (These results are only with adults.)

```{r drugs_per_visit}
f = function(design) {
  lapply(top_abx$abx, function(a) {
    svy_pm(design, as.formula(paste0(a, '~year'))) %>%
      mutate(abx=a)
  }) %>% bind_rows()
}

abx_time_model = f(design)
abx_time_model_adult = f(adult_design)
```

```{r drugs_per_visit_table}
abx_time_model_adult %>%
  filter(term=='year') %>%
  select(abx, estimate, p.value) %>%
  mutate(adj.p.value=p.adjust(p.value, method='BH')) %>%
  mutate(pct=100*(exp(estimate)-1)) %>%
  arrange(desc(pct)) %>%
  kable
```

```{r drugs_per_visit_figure}
abx_time_model_adult %>%
  filter(term=='year') %>%
  mutate(ci=1.96*std.error, y=exp(estimate)-1, ymin=exp(estimate-ci)-1, ymax=exp(estimate+ci)-1) %>%
  select(abx, y, ymin, ymax) %>%
  mutate_if(is.numeric, function(x) x*100) %>%
  arrange(y) %>% mutate(abx=forcats::fct_inorder(abx)) %>%
  ggplot(aes(x=abx, y=y, ymin=ymin, ymax=ymax)) +
  geom_point() + geom_linerange() +
  coord_flip() +
  ylab('pct change in consumption per visit') +
  theme_minimal()
```

```{r special_times}
levo_t1 = abx_time_adult %>% filter(abx=='levo', year==2011)
levo_t2 = abx_time_adult %>% filter(abx=='levo', year==2014)
amoxclav_t1 = abx_time_adult %>% filter(abx=='amoxclav', year==2011)
amoxclav_t2 = abx_time_adult %>% filter(abx=='amoxclav', year==2014)

show_var = function(df) sprintf('%g (±%g)', round(df$apc*1000, 1), round(df$se*1000, 1))
```

Levofloxacin consumption increased from `r show_var(levo_t1)` to `r show_var(levo_t2)` claims per 1,000 people per year.
Amox/clav consumption increased from `r show_var(amoxclav_t1)` to `r show_var(amoxclav_t2)` claims per 1,000 people per year.

#### Per visit, Among everyone

I do the same analysis, but among everyone.

```{r drugs_per_visit_everyone}
abx_time_model %>%
  filter(term=='year') %>%
  select(-term) %>%
  kable
```
